---
title: "Introduction to the course and the instructors"
author: "Understanding, evaluating, and improving forecasts of infectious disease burden"
format:
  revealjs:
    output: slides/introduction-to-the-course-and-the-instructors.html
    footer: "Introduction to the course and the instructors"
---

```{r instructors}
instructors <- c("sbfnk", "seabbs")
```

```{r get_contributors}
library("dplyr")
library("gh")
library("jsonlite")
whoarewe <- lapply(sort(instructors), \(instructor) {
  gh(paste0("GET /users/{username}"), username = instructor)[
    c("login", "avatar_url", "html_url", "name")
  ]
})
```

# Why this course?

::: {.incremental}
- forecasting is a common task in outbreak response
- no comprehensive training resource for infectious disease forecasting
:::

# Aim of the course

::: {.incremental}
- learn to visualize and interpret infectious disease forecasts
- evaluate forecasts using multiple approaches
- create forecast ensembles to improve predictions
:::

# Approach

Each session in the course:

::: {.incremental}
- builds on the previous one so that participants will have an overview of forecast evaluation by the end of the course
- starts with a short introductory talk
- mainly consists of interactive content that participants will work through
:::

# Who are we?

```{r whoarewe, results='asis'}
cat(":::: {.columns}\n")
for (instructor in seq_along(whoarewe)) {
  cat(
    "::: {.column width=\"", floor(100 / length(whoarewe)), "%\"}\n",
    "![](", whoarewe[[instructor]]$avatar_url, "){width=\"100%\"}\n\n",
    "[", whoarewe[[instructor]]$name, "](",
    whoarewe[[instructor]]$html_url, ")\n\n",
    ":::\n", sep = ""
  )
}
cat("::::\n")
```

# Next step

Navigate to <https://bit.ly/idm-forecasting>
